---
title: "Fire Sensitive Speleothem"
output: html_notebook
---

Following article "Past fires and post-fire impacts reconstructed from a southwest Australian stalagmite" ~ L.K. McDonough et al. 2022.

Start by loading packages and importing data (excel downloaded from https://www.sciencedirect.com/science/article/abs/pii/S0016703722001454)
Method 1) Using Excel, save data as .csv. Remove everything except column headers and data (*Note). Read as with previous data, using read_csv
Method 2) Showcased here. Tidyverse readexcel package is new to us.

```{r}
setwd("~/Dropbox/R_on_git/R_Speleothem")
library(tidyverse)
library(tidymodels)
#install.packages("readxl")
library(readxl)
#import
hi_res <- read_excel(path = "1-s2.0-S0016703722001454-mmc1.xlsx",sheet =1 )
#take a look
hi_res
```
So far looks good. Now for some QC.


```{r}
colnames(hi_res) #drop 17 and 18
hi_res <- hi_res %>% select(-(c(17,18)))
colnames(hi_res)
hi_res<-rename(hi_res, d18O="\u03b418O (\u2030)")
hi_res<-rename(hi_res, d13C="\u03b413C (\u2030)")
colnames(hi_res)
#"Al (ppm)","Ba (ppm)", "Br (ppm) smoothed using 13 point Savitsky Golay filter","Cu (ppm)",
 # "d13C","d18O", "Mg (ppm)","Organic matter (relative greyscale concentration)",
 # "P (ppm)","Pb (ppm)", "S (relative greyscale concentration)","Sr (ppm)",
 # "U (ppm)","Zn (ppm)"                   
```

Data structure (i.e., the tibble) is looking clean. Let's do some EDA (exploratory data analysis)

```{r}
summary(hi_res)
#drop columns not used in rolling average plot, and lengthen data
hi_res_long <-  hi_res %>% select(!c("Al (ppm)", "Pb (ppm)", "Cu (ppm)","Zn (ppm)"))%>%
  pivot_longer(!c("DFT (mm)","Year (CE)"),names_to = "Type",values_to = "Observation") %>% group_by(Type)

#make the new Type variable a factor (categorical) variable, and tell R what the order should be
hi_res_long$Type <- as.ordered(hi_res_long$Type)
levels(hi_res_long$Type)<- c("d13C","d18O",
 "Mg (ppm)","Organic matter (relative greyscale concentration)",
  "P (ppm)","S (relative greyscale concentration)",                           
 "Sr (ppm)","Ba (ppm)",
  "U (ppm)","Br (ppm) smoothed using 13 point Savitsky Golay filter")



#histograms of all variables
ggplot(data = hi_res_long)+geom_histogram(aes(x=Observation))+facet_wrap(Type~.,scales = "free")

#pairwise correlations
```


```{r}
#time series of all variables, 
#alternative, simpler picture for first glance: geom_smooth uses gams with a basis of penalized cubic regression splines
ggplot(data = hi_res_long)+geom_smooth(aes(x=`Year (CE)`, y=Observation))+theme_minimal()+facet_wrap(Type~.,scales = "free")
#following the paper, overlay raw data (geom_lines)
#with 5-year moving average (geom_ma)
#we could define a function to calculate moving average like this
# ma <- function(x, n = 5){stats::filter(x, base::rep(1 / n, n), sides = 2)}
#but that would be based in indices and won't account for irregular time intervals
#if i'm just looking at a single variable R base plot and hist works fine
#shows time is not regular
plot(hi_res$`Year (CE)`)
hist(hi_res$`Year (CE)`)
#if you want to know what the time steps are...
hi_res$delTime<-NA
hi_res$delTime[1]<-0
for(i in 2:length(hi_res)){
  # print(i)#reporter
  hi_res$delTime[i]<-hi_res$`Year (CE)`[i]-hi_res$`Year (CE)`[i-1]
}
plot(hi_res$delTime[2:length(hi_res$delTime)])
hist(hi_res$delTime[2:length(hi_res$delTime)],breaks = 100)

#R package for irregular time series? sure...
install.packages('zoo')
library(zoo)
is.regular(hi_res$`DFT (mm)`)
aggregate(zoo(hi_res$`DFT (mm)`))
rollmean()
ggplot(data = hi_res_long)+geom_line(aes(x=`Year (CE)`, y=Observation, color = Type))+
  geom_line(aes(x=`Year (CE)`, y=rollmean(Observation, 100, fill=NA), color =Type))+
  theme_minimal()+facet_wrap(Type~.,scales = "free",ncol = 2)+
    theme(axis.text.x = element_text(angle=45), legend.position = 'none')
p<-
  ggsave("rawFireSpeleothem.png", plot = p, device = "png", 
       scale = 1, height = 8.5, width = 11, units = c("in"),
       dpi = 300, limitsize = TRUE)
#d18O and d13C scatterplot. nonequillibrium processes at play?
ggplot(data = hi_res)+geom_point(aes(x=d18O, y=d13C))+theme_minimal()


```

